Learning how to act: making good decisions with machine learning

### 2017-05-19 thesis/offcuts.tex
draft of chapter 2 complete

\documentclass[11pt,a4paper,oneside]{book}
\usepackage[table]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algorithmic}
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{epigraph}
\usetikzlibrary{arrows,fit,positioning}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[numbers]{natbib}
\usepackage{wrapfig}

% this is included to get tables with colored rows to not draw over hlines and vlines
\usepackage{etoolbox}

\makeatletter
\patchcmd{\@classz}
  {\CT@row@color}
  {\oldCT@column@color}
  {}
  {}
\patchcmd{\@classz}
  {\CT@column@color}
  {\CT@row@color}
  {}
  {}
\patchcmd{\@classz}
  {\oldCT@column@color}
  {\CT@column@color}
  {}
  {}
\makeatother

\usepackage{amsthm}

\graphicspath{ {figures/} }
\pgfarrowsdeclarecombine{ring}{ring}{}{}{o}{o}

\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
\DeclareUnicodeCharacter{00A0}{ }

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           circle,
           rounded corners,
           draw=black, thick,
           text width=1em,
           minimum height=1em,
           text centered},
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    latentrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

\newcommand{\KL}{\operatorname{KL}}
\newcommand{\simpleregret}{R_T}
\newcommand{\Pij}[1]{\operatorname{P}_{ij}\!\left\{#1\right\}}
\newcommand{\Pkl}[1]{\operatorname{P}_{kl}\!\left\{#1\right\}}
\newcommand{\Q}[1]{\operatorname{Q}\left\{#1\right\}}
\newcommand{\EE}{\mathbb E}
\newcommand{\EEa}{\EE_a}
\newcommand{\Pns}[2]{\operatorname{P}_{#1}\left\{#2\right\}}
\newcommand{\Pn}[2]{\operatorname{P}\left\{#2|#1\right\}}
\newcommand{\actions}{\mathcal{A}}
\newcommand{\calA}{\mathcal A}
\newcommand{\etc}{\textit{etc}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\calP}{\mathcal P}
\newcommand{\x}{\boldsymbol{x}}
\newcommand{\Ps}{\operatorname{P}}
\newcommand{\R}{\mathbb R}
\newcommand{\Pri}[1]{\operatorname{P}_i\left\{#1\right\}}
\newcommand{\Prz}[1]{\operatorname{P}_0\left\{#1\right\}}

\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}[1]{\mathbb E\left[{#1}\right]}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\newcommand{\bigo}[1]{\mathcal{O}\left( #1 \right)}
\newcommand{\bigotilde}[1]{\tilde{\mathcal{O}}\left( #1 \right)}
\newcommand{\bigtheta}[1]{\Theta\left( #1 \right)}
\newcommand{\bigthetatilde}[1]{\tilde{\Theta}\left( #1 \right)}
\newcommand{\bigomega}[1]{\Omega\left( #1 \right)}

\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}
\newcommand{\cf}[2]{{#1}_{#2}}
\newcommand{\bernoulli}{\operatorname{Bernoulli}}
\newcommand{\dirac}{\operatorname{Dirac}}
\newcommand{\parents}[1]{\operatorname{\mathcal{P}a}_{#1}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}

\theoremstyle{plain} 
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\let\temp\epsilon
\let\epsilon\varepsilon

\author{Finnian Lattimore}
\title{Learning how to act: making good decisions with machine learning}

\begin{document}
\todo{figure out theoremstle is undefined}
\def\ci{\perp\!\!\!\perp} % from Wikipedia

\maketitle

\epigraph{In vain the grave, with retrospective Eye,
Would from the apparent what conclude the why,
Infer the Motive from the Deed, and show
That what we chanced, was what we meant, to do.}{Alexander Pope}

3100 words at the start of boot camp.

@@

\chapter{Introduction}

\paragraph*{My thesis in a sentence:} Unifying causal inference with multi-armed bandits.

\paragraph*{This thesis contributes to knowledge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.

\paragraph*{My key research questions are:} 
\begin{itemize}
\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
\item To summarise the key strands of causal inference research from economics and social sciences for the machine learning community
\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
\end{itemize}

\section{Motivation}
Many of the most important questions in science and in our personal lives are about the outcomes of doing something. Will asking people to pay upfront at the doctors reduce long term health expenditure? If we developed a drug to suppress particular genes, could we cure MS and would delaying teen-aged pregnancies improve the outcome for their kids.

These are hard questions because they require more than identifying a pattern in data. Correlation is not causation. Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, where the variables involved are susceptible to human intuition and understanding.

We now live in a world of data. Hours of our lives are spent online, where every click can be recorded, tiny computers and sensors are cheap enough to incorporate into everything and the US Institute of Health is considering if all infants should be genetically sequenced at birth. Such data gives us a window into many aspects of our lives at an unprecedented scale and detail but it is messy, complicated and often generated as a by product of some other purpose. It does not come from the controlled world of a randomised experiment.

The rise of big data sets and powerful computers has seen an explosion in the application of machine learning. From healthcare, to entertainment and self driving cars; machine learning algorithms will transform many industries. It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artificial intelligence exceeds our own and then grows exponentially.

However, despite the huge advances in specific areas of machine learning (in particular deep learning), machine learning algorithms are effective only within very narrow problem settings. Getting them to generalise to even slightly different problems or data sets remains very challenging. Deciding how we should act or what policies we should implement requires us to be able to predict how a system will behave if we change it. The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on. As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live it, it is critical to ensure that we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.

\section{What is causality and why do we care? (2000 words)}

The notion of causality has been widely debated in science and philosophy \cite{} but is still widely viewed as poorly defined \cite{}. This has led to a reluctance among applied researchers in many fields to refer to cauality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existance of an association depends on what other variables we adjust for (or include in a regression). Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense if the association reported is the \emph{right one}.

We discuss more detailed definitions of causality in section c

The what-if type questions from the why. \cite{Gelman2010} \cite{}. Why do whites to better than blacks in school (in America). Suggests that reverse causal inference questions are more interesting and motivate most of social science.

I do not find this distinction useful. 
We can only change the future - history is useful only as far as it tells us something about the future.
Reverse causal questions can be reposed as forward ones, when making the translation a reverse causal query will be effectively asking about many possible interventions (rather than just one). Problems highlighted as intractible in the reverse causal sense are also intractible in the forward inference form, typically because concern situations for which we do not have a sufficient number of similar instances to allow statistical reasoning. For example, the war question posed in Gelman.

A distinction between forward causal iference, what happens if we do X and reverse casual inference

To explain or to predict \cite{Shmueli2010}
The two cultures \cite{Breiman2001}

\paragraph*{There are two reasons why correlation is not causation}. The first is related to variance and over-fitting. Observations are noisy. With a finite data set with enough variables we will be able to find some that are completely unrelated but correlate purely by chance. \ref{figure - search for random correlation}. The second arises from bias, typically introduced by an un-observed confounding variable. In this case, variables are correlated not by chance. We would expect the relationship to hold if we sampled more data. However, they are not causally related in that intervening to set one would not likely effect the other. EXAMPLE WITH FIGURE.

\subsection{Defining causality}
\begin{itemize}
\item widely debated in science and philosophy (REFERENCES)
\item what is explanation?
\item any model that aims to predict the outcome of an action or intervention in a system
\item I do not see the distinction between explanation and (causal) prediction. Explanation is all about the ability to compress and to generalise. The more a model can do this, the more we view as providing an understanding of the why. 
\item mediation?
\end{itemize}
\subsection{Overview of this thesis}
\subsection{Identifying when we have a causal problem}

\subsubsection{Examples of typical machine learning problems. Are they causal?}

Consider the following problems, which span a wide range of the types of questions machine learning is currently being applied to. Which of them require casual inference? How can we identify characteristics of a problem that make causal modelling important?

\begin{itemize}
\item Speech recognition (for systems like Siri or Google)
\item Machine translation 
\item Image classification
\item Forecasting the weather
\item Playing Go 
\item Identifying spam emails
\item Automated essay marking
\item Predicting the risk of death in patients with pneumonia.
\item Predicting who will re-offend on release from prison 
\item Predicting which customers will cease to be your customers
\item Demand prediction for inventory control
\item Predicting who will click on an ad
\item Financial trading
\item Recommending movies
\item Online search
\item Self driving cars
\item Pricing insurance
\end{itemize}

The above problems are not posed with enough detail to know if causality is an important consideration. In particular, I failed to specify what actions the might be taken in response to model.

Consider speech recognition. You say something, which causes to sound waves, which are converted to a digital signal which Siri maps to words. Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a DVD on elocution). A similar argument could be made for many applications of machine translation and image classification.

In image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see. We can be fairly confident that the process will not change. If we develop a discriminating model that is highly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).

What about forecasting the weather? If you are using a short term forecast to decide whether to pack an umbrella it's clear causality can be ignored - your decision will not effect if it actually rains. However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system. For this we need a (causal) model that allows us to predict the outcome under various different interventions.

Identifying spam and automated essay marking systems are similar. The decision made by the algorithm is likely to change the relationship between the features used by the algorithm and the true label. Spammers and students will modify their writing in order to optimise their results. The standard machine learning approach can only work if the resulting change is sufficiently gradual and fresh ground truth (probably human labelled) training data is provided. (What would the nature of the features have to be such that change did not occur? - they would have to be causes of the label).

What about predicting the risk of death in patients with pneumonia? Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective this seems like a straightforward prediction problem. It is not. Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if we start using the model to decide whom to admit. (xxx et al) found exactly this effect. Their model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pneumonia. They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk. There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addmition and treatment protocols. However, using it to decide on what basis to admit people could kill. The key is understanding exactly what question you are asking. In this case we are care about what happens to patients with characteristics X if we treat them according to decision rule Z.

Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (loyalty reward, deny parole or more support whilst on parole, etc). For all these problems the assumptions required to treat them as pure prediction problems are;
\begin{enumerate}
\item The treatment is assumed to be effective (at least better than nothing)
\item Deciding who to treat on the based of the model predictions won't change the relationship between features and outcome 
\end{enumerate}

Demand prediction seems relatively straightforward. These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product. It seems unlikely that using the model to ensure stock is available will itself change demand. However there is a potential data censoring issue. If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero. Changing availability does then change demand.

Playing Go (and other games) is another case with some subtleties. At every turn, the AI agent has a number of actions available. The board state following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win. ... this is interesting maybe come back to it ... an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ... One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken). When can we learn from the actions others have taken? When there is no confounding. And does this hold with Go? Probably because the board state encapsulates everything that should determine what move is played. Learning directly from actions and trying to generalise (can in some instances reduce the problem to standard ML)

Having considered these examples we can now identify some general aspects of the problem that determine whether or not we require a causal model.

\begin{itemize}
\item Does acting on the predictions of the model change the mapping from features to target? (at least if the decision process is open to scrutiny). In general, if we believe that humans are generally trying to optimise to various goals of their own then for any system interacting with them the answer to this will be yes.
\item

 Covariate shift clearly comes in here. Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occurring and that results will be transferable. The mechanism is known but the function may be complex. Can we write down something that causal models are invariant to in terms of shift that is not the case for non-causal models? Yes, if the way in which features get their values changes, then causal models will be invariant to that in a way that non-causal ones are not.

\item To decide between actions we only need to rank them (not estimate their actual effect). 
\item The predicted outcome in the absence of an intervention provides a single point. We can use this to find which problems are most serious if left alone - and prioritise those for modelling changes. 
\item Any decision we take does not significantly impact the system from which the data was drawn to make it (for repeat decision making)
\item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
\item Ethics - ... People's viewpoint on if its OK...
\end{itemize}

I hope these examples gave you a feel for the richness and subtitles of causal inference. We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.

\section{Approaches to causality (1000 words)}

There are two broad approaches to deciding how to act. Reinforcement learning and causal inference. In reinforcement learning we estimate the effect of actions by taking them. We assume there is an agent capable of intervening in the system and try to find policies for the agent to follow in selecting actions so as to maximise some kind of reward. This is a very powerful and general framework (REFERENCES TO GENERAL AI). However, POINT OUT SOME OF THE DIFFICULTIES WITH REINFORCEMENT LEARNING. We are frequently presented with large bodies of data that have been collected from a system in which we did not have any control over what actions were taken.

We will call data sets where we do not have control over the decision making process that generated the data observational. Versus experimental data sets, where we do have control (experimental data sets do not always have to be randomised -although that is a powerful approach to ensure we have control. There is a space in the middle where we have partially controlled the process by which agent select actions. Randomised data with imperfect compliance would be an example.

An agent (capable of intervening in the system) chooses an action from those available The agent making the decision included in the model. This is

Reinforcement learning addresses the problem of learning from explicitly taking actions. There
is typically some state or environment. An agent chooses an action from those available in the
current state. The state then evolves stochastically as a function of the selected action and
the agent receives some feedback or reward that is a function of the new state. This setting
differs from the standard classification problem in that the agent must learn from feedback on
the selected action, rather than being presented with the correct action for a given state. A
common modelling assumption is that the state evolves only as a function of the previous state
and the action chosen, and given these, is independent of the previous history of states and
actions. This is known as a Markov decision process or MDP. A particularly well studied and
understood model is the singe state MDP. In this case, there are a set of actions, each associated
with a fixed but unknown reward distribution and at each time step our agent selects an action
and receives corresponding feedback. This is known as the multi-armed bandit problem.

Causal inference makes use of assumptions to allow the outcome of actions to be predicted from
observational data. The key to causal inference is a framework that can model how actions
change the state of the world. This framework then allows us to map information collected in
one setting to another.

Both approaches can be seen as extensions to the concept of randomised controlled trials. Bandit
algorithms deal with the sequential nature of the decision making process, causal inference with
the problem that full randomisation is not always feasible, affordable or ethical. The similarities
between the problems that these techniques have been developed to address raises the question
of if there are problems best addressed by a combination of these approaches and how they
can be combined. The goal of my thesis is to explore these questions. In the next sections I
review the key literature in causal inference and bandits. I then present a general approach to
how causal models might be incorporated into bandit settings and conclude by demonstrating
an algorithm that leverages causal assumptions to improve performance in a specific bandit
setting.

There are two key approaches to causal problems. The first is to learn the outcome of actions by directly intervening in the system and seeing what happens. We then get feedback on how good those actions were. This is the approach taken in reinforcement learning. ADVANTAGES AND DISADVANTAGES OF THIS APPROACH. The second broad approach is causal inference. Here

\todo{insert figure showing data generating process and observed data defining ML}\ref{all of statistics}

\paragraph*{Two broad approaches} 
\begin{itemize}
\item Build a model to map the natural behaviour of the system to what will happen for some action
\item Take the action and see what happens
\end{itemize}

\paragraph*{The first is causal inference}

\paragraph*{The second is reinforcement learning}

\paragraph*{Both generalise from randomised experiment} Reinforcement learning to sequential decisions, causal inference to non-experimental conditions

Both these fields relate to the problem of making optimal decisions and both can be seen as generalising randomised controlled experiments. Causal inference  is the study of how to estimate the effect of an action in the absence of randomisation. Reinforcement learning studies how we can do better if the decisions are to be made sequentially.

\paragraph*{Both approaches involve assumptions} the latter that we can group context and actions.

\paragraph*{Limitations of causal inference}

\paragraph*{Limitations of experiments} What are the issues with standard randomised experiments?

\chapter{Causal models}
\label{chap:causal_models}

Causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it. To do this we need terminology to describe actions and how we anticipate the system should respond to them. Three key approaches have emerged; counterfactuals, structural equation models and causal bayesian networks. In this chapter we describe these approaches. We will examine what problems they allow us to solve, what assumptions they rely on and how they differ. We will also use them to describe the following simplified examples. The aim is to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.

\vspace*{.3cm}
\begin{example}
\label{exm:ranomized_experiment}
Suppose we have developed a new drug for some illness and wish to determine how effective it is. We take a large group of patients and randomly assign half of them to a treatment group and the other half to a control group. The people in the treatment group get the drug, everyone else gets a placebo pill. The question we want to answer is does giving people the active drug improve their changes of recovery relative to giving them the placebo. We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome. 
\end{example}

\vspace*{.3cm}
\begin{example}
\label{exm:adjusting}
Suppose we wish to estimate what would happen to high school graduation rates if we made pre-school available and compulsory for all 4 year olds. We have a large cross-secional dataset on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES). We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.  
\end{example}

\section{Causal bayesian networks}

Causal Bayesian networks are an extension of Bayesian networks. A Bayesian network is a graphical way of representing how a distribution factorises. Any joint probability distribution can be factorised into a product of conditional probabilities. There are multiple valid factorisations, corresponding to permutations of variable ordering.

\eqn{
\label{eqn:cbn:joint_dist}
P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
}

We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution, see figure \ref{fig:bayesnet}. If the factorisation simplifies due to conditional independencies between variables, this is reflected by missing edges in the corresponding network. There are multiple valid Bayesian network representations for any probability distribution over more than one variable, see figure \ref{fig:bayesnet2} for an example.

\begin{figure}[h]
\centering
\caption{A general Bayesian network for the joint distribution over three variables. This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
\label{fig:bayesnet}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below left=of 1](2){$X_{2}$};
\node[main node, below right=of 1](3){$X_{3}$};

\begin{figure}
\centering
\caption{Some valid Bayesian networks for a distribution that an be factorised as $\P{X_1,X_2,X_3} = \P{X_1}\P{X_2}\P{X_3|X_2}$ (which implies $X_3 \ci X_1 | X_2$)}
\label{fig:bayesnet2}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.15\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
 %nodes
\node[main node](1){$X_{1}$};
\node[main node, below=of 1](2){$X_{2}$};
\node[main node, below=of 2](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (3) edge (2)
    (2) edge (1);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, below left=of 1](2){$X_{1}$};
\node[main node, below right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm, thick,main node/.style={punkt}]
 %nodes
\node[main node](1){$X_{2}$};
\node[main node, above left=of 1](2){$X_{1}$};
\node[main node, above right=of 1](3){$X_{3}$};
\path[every node/.style={font=\sffamily\small}]
    (2) edge (1)
    (3) edge (1)
    (2) edge (3);
\end{tikzpicture}
\end{subfigure}
\end{figure}

The statement that a given graph $G$ is a Bayesian network for a distribution $P$ tells us that the distribution can be factorised over the nodes and edges in the graph. There can be no missing edges in $G$ that do not correspond to conditional independencies in $P$ (the converse is not true $G$ can have extra edges). If we let $parents_{X_{i}}$ represent the set of variables that are parents of the variable $X_{i}$ in $G$ then we can write the joint distribution as;

\begin{equation}
P(X_{1}...X_{N}) = \prod_{i = 1...N}P(X_{i}|parents_{X_{i}})
\end{equation}

A causal Bayesian network is a Bayesian network in which it a link $X_{i} \rightarrow X_{j}$, by definition, implies $X_{i}$ causes $X_{j}$. This means that if we intervene and change the value of $X_{i}$, we expect $X_{j}$ to change, but if we intervene to change $X_{j}$, $X_{i}$ will not change. We need some notation to describe interventions and represent distributions over variables in the network after an intervention. In this thesis I use the do operator introduced by Pearl \citep{Pearl2000}.

\begin{definition}{The do-notation}
\begin{itemize}
\item do(X=x) denotes an intervention that sets the random variable(s) $X$ to $x$.
\item $\P{Y|do(X)}$ is the distribution of $Y$ conditional on an \emph{intervention} that sets $X$. This notation is somewhat overloaded. It may be used represent a probability, a probability distribution/mass function or a family of distribution functions depending on if the variables are discrete or continuous and whether or not we are treating them as fixed. For example it could represent 
\begin{itemize}
\item The probability $\P{Y=1|do(X=x)}$ as a function of $x$
\item The probability mass function for a discrete $Y$ : $\P{Y|do(X=x)}$
\item The probability density function for a continuous  $Y$ : $f_Y(y|do(X=x))$
\item a familiarly of density/mass function for $Y$ paramaterised by $x$.
\end{itemize}
Where the distinction is important not clear from context we will use one of the more specific forms above. 
\end{itemize}
\end{definition}

\begin{theorem}[Truncated product formula]
\label{thm:truncated_prodcut}
If $G$ is a causal network for a distribution $P$ defined over variables $X_{1}...X_{N}$, then we can calculate the distribution after an intervention where we set $Z \subset X$ to $z$, denoted $do(Z=z)$ by simply dropping the terms for each of the variables in $Z$ from the factorisation given by the network \citep{Pearl2000}.

\begin{equation}
\label{eq:truncatedproduct}
\P{X_1...X_N|do(Z=z)} =
  \begin{cases}
  \prod_{i \notin Z}\P{X_{i}|parents_{X_{i}}} & \text{if $(X_{1}...X_{N})$ consistant with $Z=z$}  \\
   0       & \text{otherwise } 
  \end{cases}
\end{equation}
\end{theorem}

Theorem \ref{thm:truncated_prodcut} does not hold for standard bayesian networks for the simple reason that there are multiple valid networks for the same distribution. The truncated product formula will give different results depending on which you choose. The result is possible with causal bayesian networks because it follows directly from the assumption that the direction of the link indicates causality. In fact, from the interventionist viewpoint of causality, the truncation product formula defines what it means for a link to be causal.

Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$? The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and disease sub type of the patient. Since $X$ is assigned  via deliberate randomisation we know that it is not effected by any latent variables. The causal bayesian network for this scenario is shown in figure \ref{fig:causal_network_example} .This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$. In this example, the interventional distribution is equivalent to the observational one.

\begin{figure}[h]
\caption{Causal Bayesian network for example \ref{exm:ranomized_experiment}}
\label{fig:causal_network_example}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){$X \text{ (Treatment)}$};
\node[node, right=of 2](3){$Y \text{ (Outcome)}$};
\path[]
	(2) edge (3);
\end{tikzpicture}
\end{figure}

In example \ref{exm:adjusting} we are interested $\P{Y|do(X=1)}$, the expected high-school graduation rate if we introduce universal preschool. We could compare it to outlawing pre-school $\P{Y|do(X=0)}$ or the current status quo $\P{Y}$. It seems reasonalble to assume that attending could pre-school effect the likelyhood of graduating from high school and that your parents socio-economic status effects both pre-school attendance and high-school graduation. If we assume that socio-economic status is the only such variable (nothing else effects both attendance \emph{and} graduation), we can represent this problem with the causal bayesian network in figure \ref{exm:adjusting}. In this case, the interventional distribution is not equivelent to the observational one. If better off parents are more likely to send their children to pre-school these children more likely to graduate high school regardless, then simply comparing the high school graduation rates of those who attended preschool with those who did not will overstate the benifit of preschool. To obtain the interventional distribution we have to estimate the benifit of preschool for each socio-economic level seperately and then weight the results by the proportion of the population in that group, $\P{Y|do(X=1)} = \sum_{z \in Z}\P{Y|X=1,Z}\P{Z}$.

We have seen from these two examples that the expression to estimate the causal effect of an intervention depends on the structure of the causal graph. There is a very powerful and general set of rules that specify how we can transform observational distributions into intervetional ones for a given graph structure. These rules are referred to as the Do-calculus \cite{Pearl2000}. We discuss them further in section \ref{}.

\begin{figure}
\center
\caption{Causal bayesian network for example \ref{exm:adjusting}}
\label{fig:causal_adjust}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.2cm, thick,node/.style={observedrect},lt/.style={latent}]

 %nodes
\node[node](1){$Z \text{ (SES)}$};
\node[node, below left=of 1](2){$X \text{ (Pre-school)}$};
\node[node, below right=of 1](3){$Y \text{ (Graduated)}$};
\path[every node/.style={font=\sffamily\small}]
    (1) edge (2) edge (3)
    (2) edge (3); 
\end{tikzpicture} 
\end{figure}

Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention. The power of defining an intervention in this way stems from the number of things that are invariant between the two processes. All the (conditional) distriubtions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.

We could represent problems of the type where we try to infer properties of the post-interventional system based on data generated by the pre-interventional distribution by explicitly representing both systems and what they have in common, see figure \ref{fig:causal_ordinary_bayes}. This does not require any special framework or notation. The graphs in figure  \ref{fig:causal_ordinary_bayes} are ordinary Bayesian networks. However, without a causal framework, we have to make assumptions about what will be invarient to the intervention specifically for each such problem we encounter. For complex problems, it is very difficult to conceptualise what assumptions we expect to hold without the benifit of a causal framework.

>>>>

\section{Clarifying some confusing definitions}

With a causal framework in place, we are now in a position to clarify some confusing concepts that are very prevelent in economics and social sciences.

Confounding
Exogenous
Endogenous
Selection bias
...

@@
